---
title: "Tidy Data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

<!-- This file by Jake Thompson is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup, include = FALSE}
library(tidyverse)

# Toy data
cases <- tribble(
  ~country, ~"2011", ~"2012", ~"2013",
      "FR",    7000,    6900,    7000,
      "DE",    5800,    6000,    6200,
      "US",   15000,   14000,   13000
)

pollution <- tribble(
       ~city,   ~size, ~amount,
  "New York", "large",      23,
  "New York", "small",      14,
    "London", "large",      22,
    "London", "small",      16,
   "Beijing", "large",     121,
   "Beijing", "small",      56
)

scores <- tribble(
                     ~name,       ~house, ~score,
          "Ronald Weasley", "Gryffindor",     78,
            "Harry Potter", "Gryffindor",     85,
        "Hermione Granger", "Gryffindor",    100,
  "Justin Finch-Fletchley", "Hufflepuff",     87,
            "Hannah Abbot", "Hufflepuff",     92,
             "Susan Bones", "Hufflepuff",     93,
       "Anthony Goldstein", "Ravenclaw",      84,
          "Michael Corner", "Ravenclaw",      93,
             "Padma Patil", "Ravenclaw",      97,
          "Vincent Crabbe", "Slytherin",      61,
           "Gregory Goyle", "Slytherin",      61,
            "Draco Malfoy", "Slytherin",      92
)


bp_systolic <- tribble(
  ~ subject_id,  ~ time_1, ~ time_2, ~ time_3,
             1,       120,      118,      121,
             2,       125,      131,       NA,
             3,       141,       NA,       NA 
)

bp_systolic2 <- tribble(
  ~ subject_id,  ~ time, ~ systolic,
             1,       1,        120,
             1,       2,        118,
             1,       3,        121,
             2,       1,        125,
             2,       2,        131,
             3,       1,        141
)
```


## Tidy and untidy data

`table1` is tidy:

```{r}
table1 
```

`table2` isn't tidy, the count column really contains two variables:

```{r}
table2
```


## Your Turn 1

Is `bp_systolic` tidy?

```{r}
bp_systolic
```


## Your Turn 2

Using `bp_systolic2` with `group_by()`, and `summarize()`:

* Find the average systolic blood pressure for each subject
* Find the last time each subject was measured

```{r}
bp_systolic2
```


## Your Turn 3

On a sheet of paper, draw how the cases data set would look if it had the same values grouped into three columns: **country**, **year**, **n**


## Your Turn 4

Use `pivot_longer()` to reorganize `table4a` into three columns: **country**, **year**, and **cases**.

```{r}
table4a 
```


## Your Turn 5

On a sheet of paper, draw how `pollution` would look if it had the same values grouped into three columns: **city**, **large**, **small**


## Your Turn 6

Use `pivot_wider()` to reorganize `table2` into four columns: **country**, **year**, **cases**, and **population**.

```{r}
table2 
```


## separate() and unite()

```{r}
scores %>%
  separate(name, into = c("first", "last"), sep = " ")
```

```{r}
sep_scores <- scores %>%
  separate(name, into = c("first", "last"), sep = " ")
sep_scores %>%
  unite("full_name", first, last, sep = " ")
```


***

# Take Aways

Data comes in many formats but R prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column
2. Every observation is in its own row
3. Every value is in its own cell (which follows from the above)

What is a variable and an observation may depend on your immediate goal.
