---
title: "Organize with List Columns"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

<!-- This file by Jake Thompson is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup, include = FALSE}
library(tidyverse)
library(broom)
library(modelr)
library(rsample)
library(gapminder)
library(tidydscompanion)
library(here)
```


## Your Turn 1

How has life expectancy changed over time?

* Make a line plot of lifeExp vs. year grouped by country.  
* Set alpha to 0.2, to see the results better.

```{r}
gapminder
```


## Quiz

How is a data frame/tibble similar to a list?


## Quiz

If one of the elements of a list can be another list, can one of the columns of a data frame be another list?


## Your Turn 2

Run this chunk:

```{r}
gapminder_nested <- gapminder %>%
  group_by(country) %>%
  nest()

fit_model <- function(df) lm(lifeExp ~ year, data = df)

gapminder_nested <- gapminder_nested %>% 
  mutate(model = map(data, fit_model))

get_rsq <- function(mod) glance(mod)$r.squared

gapminder_nested <- gapminder_nested %>% 
  mutate(r.squared = map_dbl(model, get_rsq))
```

Then filter `gapminder_nested` to find the countries with r.squared less than 0.5.  

```{r}

```


## Your Turn 3

Edit the code in the chunk provided to instead find and plot countries with a slope above 0.6 years/year.

```{r}
get_slope <- function(mod) {
  tidy(mod) %>% filter(term == "year") %>% pull(estimate)
}

# Add new column with r-sqaured
gapminder_nested <- gapminder_nested %>% 
  mutate(r.squared = map_dbl(model, get_rsq))

# filter big slope countries
poor_fit <- gapminder_nested %>% 
  filter(r.squared < 0.5)

# unnest and plot result
unnest(poor_fit, data) %>%
  ggplot(aes(x = year, y = lifeExp)) + 
    geom_line(aes(color = country))
```


## Your Turn 4

**Challenge:**

1. Create your own copy of `gapminder_nested` and then add one more list column: `output` which contains the output of `augment()` for each model.
2. Plot the residulals against time for the countries with small r-squared.

```{r}

```


## Bootstrapping Comparisons

```{r}
mean(admission$gre_v[admission$gender == "Male"]) -
  mean(admission$gre_v[admission$gender == "Female"])

mean_diff <- function(splits) {
  x <- analysis(splits)
  mean(x$gre_v[x$gender == "Male"]) -
    mean(x$gre_v[x$gender == "Female"])
}

set.seed(32011)
grev_gender <- admission %>%
  bootstraps(times = 100) %>%
  mutate(grev_diff = map_dbl(splits, mean_diff))

ggplot(grev_gender, mapping = aes(x = grev_diff)) +
  geom_density()

quantile(grev_gender$grev_diff, probs = c(0.025, 0.500, 0.975))
```


## Your Turn 5

Is there a difference between the percentage of male and female applicants admitted?

First, write some code to calculate the difference between the percentage of male applicants admitted, and the percentage of female applicants admitted.

```{r}

```

Now, turn that code into a function, `pct_diff()`.

```{r}
pct_diff <- function(splits) {
  x <- analysis(splits)
  
}
```

Generate 100 bootstraps, and apply the `pct_diff()` function to each bootstrapped sample.

```{r}
set.seed(32011)
admit_gender <- admission %>%
  ___ %>%
  mutate(admit_diff = ___)
```

Finally, create a density plot of the difference, and create a 95% interval.

```{r}
ggplot(admit_gender, mapping = aes(___)) +
  ___()

quantile(___, probs = c(0.025, 0.500, 0.975))
```


***

# Take away

* Store objects and other lists in list-columns of data frames
* Use `bootsraps()` to recreate resampled data objects
* Use `vfold_cv()` to create analysis and assessment sub-samples of your data to assessment model performance
* Use `purrr` to iterate over bootstrapped samples and cross validation folds
