---
title: "Model Data - Solutions"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

<!-- This file by Jake Thompson is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup, include = FALSE}
library(tidyverse)
library(modelr)
library(broom)

wages <- heights %>% filter(income > 0)
```


## Your Turn 1

Fit the model on the slide and then examine the output. What does it look like?

```{r}
mod_e <- lm(log(income) ~ education, data = wages)
mod_e
```


## Your Turn 2

Use a pipe to model `log(income)` against `height`. Then use broom and dplyr functions to extract:

1. The **coefficient estimates** and their related statistics 
2. The **adj.r.squared** and **p.value** for the overall model

```{r}
mod_h <- wages %>% lm(log(income) ~ height, data = .)
mod_h %>%
  tidy()

mod_h %>% 
  glance() %>% 
  select(adj.r.squared, p.value)
```


## Your Turn 3

Model `log(income)` against `education` _and_ `height`. Do the coefficients change?

```{r}
mod_eh <- wages %>% 
  lm(log(income) ~ education + height, data = .)

mod_eh %>% 
  tidy()
```


## Your Turn 4

Model `log(income)` against `education` and `height` and `sex`. Can you interpret the coefficients?

```{r}
mod_ehs <- wages %>% 
  lm(log(income) ~ education + height + sex, data = .)

mod_ehs %>% 
  tidy()
```


## Your Turn 5

Use a broom function and ggplot2 to make a line graph of `height` vs `.fitted` for our heights model, `mod_h`.

_Bonus: Overlay the plot on the original data points._

```{r}
mod_h %>% 
  augment(data = wages) %>% 
  ggplot(mapping = aes(x = height, y = .fitted)) +
    geom_point(mapping = aes(y = log(income)), alpha = 0.1) +
    geom_line(color = "blue")
```


## Your Turn 6

Repeat the process to make a line graph of `height` vs `.fitted` colored by `sex` for model mod_ehs. Are the results interpretable? Add `+ facet_wrap(~education)` to the end of your code. What happens?

```{r}
mod_ehs %>% 
  augment(data = wages) %>% 
  ggplot(mapping = aes(x = height, y = .fitted, color = sex)) +
    geom_line() +
    facet_wrap(vars(education))
```


## Your Turn 7

Use one of `spread_predictions()` or `gather_predictions()` to make a line graph of `height` vs `pred` colored by `model` for each of mod_h, mod_eh, and mod_ehs. Are the results interpretable? 

Add `+ facet_grid(sex ~ education)` to the end of your code. What happens?

```{r}
wages %>% 
  gather_predictions(mod_h, mod_eh, mod_ehs) %>%
  filter(education > 11) %>%
  ggplot(mapping = aes(x = height, y = pred, color = sex)) +
    geom_line() +
    facet_grid(rows = vars(model), cols = vars(education))
```


##  Your Turn 8

Use one of `spread_residuals()` or `gather_residuals()` to make a scatter plot of `afqt` vs. `resid` for each of mod_e, mod_h, mod_eh, and mod_ehs.

Use a faceting function to create a subplot for each model.

```{r}
wages %>%
  gather_residuals(mod_e, mod_eh, mod_ehs, mod_h) %>%
  ggplot(mapping = aes(x = afqt, y = resid)) +
    geom_point() +
    facet_wrap(vars(model))
```

```{r}
wages %>%
  gather_residuals(mod_e, mod_eh, mod_ehs, mod_h) %>%
  ggplot(mapping = aes(x = afqt, y = resid)) +
    geom_hex() +
    scale_fill_viridis_c() +
    facet_wrap(vars(model))
```


***

# Take Aways

* Use `glance()`, `tidy()`, and `augment()` from the **broom** package to return model values in a data frame.

* Use `add_predictions()` or `spread_predictions()` or `gather_predictions()` from the **modelr** package to visualize predictions.

* Use `add_residuals()` or `spread_residuals()` or `gather_residuals()` from the **modelr** package to visualize residuals.
